---
title: "Exploratory Analysis"
output: html_document
---

## Project title

Shooting for Glory: A Data-Driven Exploration of Clutch Performance in the NBA

### Authors

Isaac Ramirez

### Date

Spring 2024

### Abstract

Our project explores the concept of clutch shooting in the NBA, analyzing the
performance of players in high-pressure situations. This question is
important because it can provide insights into the mental and physical
preparation required for success in critical game moments. To address this
question, we will examine the relationship between clutch shooting
percentages and other factors such as game experience, shot difficulty, and
player consistency.

### Keywords

NBA, Clutch Shooting, Basketball Analytics, Free throw, shot difficulty

### Introduction  

The capacity to perform well under pressure is a highly valued quality in
sports. This is also called clutching. Additionally, clutch shooting is the
ability to make important shots at game-changing moments. Many shots have
transformed many NBA and WNBA games, series, and seasons. This research
examines 1996-97 NBA clutch shooting statistics. Physical skill, mental
fortitude, and situational awareness are needed for clutch shooting. Players
with clutch genes can focus, adapt, and shoot accurately and confidently. The
importance of clutch shooting cannot be overstated, as it can swing the
momentum of a game and ultimately decide the outcome. Below are the Research
questions: How do clutch shooting percentages vary by game experience? How
much of an effect does shot difficulty relate to the outcomes of clutch
shooting percentages? Do clutch shooters repeat this ability to make clutch
shots consistently? These questions may show mental and physical preparedness
for high-pressure success. Patterns and trends in the data may affect
coaching and player development. Clubs may offer veteran players additional
opportunities at important times if they perform well. If clutch shooting and
shot difficulty are associated, athletes who can make tough shots under
pressure may thrive. We hope to highlight clutch shooting's complex mechanics
and basketball impact by addressing these questions.

### Related Work  

Numerous factors impact basketball players' clutch performance, which has
been extensively studied. Considerations include teamwork, player experience,
and game setting. In stressful situations, players who can stay cool and
focused are more likely to succeed, according to research. Analytics like win
probability and clutch scoring percentages have identified NBA clutch
performers. Beyond points per game and shooting percentages, these
measurements show a player's impact. Also studied are clutch shot mechanics
and accuracy. Stephen Curry and other players with a longer shot arc are more
accurate in crucial situations, according to research. This stresses
technical skill in clutch performance and claims that players who can make
challenging shots would perform better under pressure. WNBA clutch players
like Arike Ogunbowale, Courtney Vandersloot, and Brittney Griner thrive under
pressure. Their unique mix of physical skill, mental grit, and situational
awareness helps them succeed in vital situations. Basketball clutch
performance research illustrates how physical and emotional factors combine
to influence success under pressure. Analyzing these factors and measuring
clutch performance can help us understand what makes a player clutch and how
teams should strategy to win key games. All of my insights from above are
laid out in the references of relevant work below. References: https://
gamerant.com/nba-2k22-best-wnba-player-each-position-rating-pg-sg-sf-pf-c/
https://en.wikipedia.org/wiki/Alysha_Clark https://www.inpredictable.com/p/
about-this-site.html?m=1 https://www.nba.com/news/stats-leaders-who-are-top-
clutch-scorers-this-season https://www.inpredictable.com/2015/05/introducing-
sharc-shot-arc-analysis.html?m=1

### The Dataset

The Dataset The nba.csv file includes all NBA clutch shot takers since 1996-
97. Data from a list of prominent clutch shots (~64,000) contains player IDs,
names, clutch free throw attempts, total clutch shots, clutch shooting
percentages, and more. Data Source The data was gathered and given by
inpredictable, a company that provides insights on many sports, including
basketball. Data Gathering Data was acquired manually using
basketballreference.com, assuring correctness and dependability. Data on
point difference at the 5-minute mark, when clutch time begins, was manually
collected using basketballreference.com. The data also contains manually
gathered statistics like minutes played, field goals attempted, and free
throws taken up until clutch time. Data Collection Goal The study examined
clutchness in basketball, specifically late-game free throw shooting in
playoffs and the impact of point disparity. Features of the dataset The
dataset has 200 observations (players) and 9 characteristics (columns),
making it acceptable for analysis. Combining basic data like rebounds with
sophisticated metrics like pace or use rate enables a comprehensive analysis
of crucial performance. Moral Issues Consider ethical problems and power
dynamics when handling this data, such as possible bias in data collecting
and processing, and the effect of results on players and teams. The dataset's
concentration on clutch time free throw % as the dependent variable may
overemphasize one element of clutch performance, omitting other crucial
aspects

### Implications

There are many implications that could be affected by my analysis of players
that are clutch. For example, Identifying clutch shooting performance markers
aids GM team-building decisions. To make the optimal decisions as a gm you
need to recruit high-pressure players, developing clutch shooting training
programs, and advising on game strategy and player selection. To clarify, a
general manager recruits and signs the top players. Players that perform well
under pressure are essential to winning late in the games. Data may help GMs
evaluate player clutch characteristics and make informed roster changes. This
may include signing clutch shooters or free agents with the skills and
perseverance to succeed under pressure. This study may affect GM contract
talks and player evaluations. Understanding clutch shooting's importance may
help GMs negotiate contracts and evaluate players. Teams who can attract key
clutch players may have a superior roster and competitive advantage. The NBA
and WNBA may adjust team construction and administration according to the
findings of my analysis.

### Limitations & Challenges
The findings of "Shooting for Glory: A Data-Driven Exploration of Clutch
Performance in the NBA," must overcome numerous challenges to be reliable.
Three important concerns need significant consideration. First Challenge:
Data Limits Although vast, the dataset may not capture all clutch shooting
instances. It emphasizes playoff late-game free throw shooting, which may not
reflect other crucial moments. The dataset may not contain important regular-
season or non-playoff shots. This restricted view may not explain clutch
performance since players act differently in different games. Shot selection,
defense, and collaboration may be overlooked in the dataset's crunch time
free throw % concentration. No one statistic can measure clutch performance.
The study solely examined free throw %, which may have missed clutch shooting
nuances. Problem 2: Methodology To compare clutch shooting percentages to
other criteria, use statistical models. These models may not accurately
represent clutch performance due to team chemistry, coaching, and player
psychology. Clutch performance may rely on teammates, coach, and mental
preparation. Intangibles may lead statistical models to oversimplify variable
relationships. Statistical models may overemphasize quantitative measures,
concealing qualitative findings. Intangible traits that set great athletes
apart also impact clutch success. Over-statistical modeling may impede clutch
performance studies from comprehending humans. 3. Interpretation Issues This
study is interpretable, and "clutch" is subjective. What is a clutch
situation, and how do we account for player anxiety in various games? Are
playoff game-tying shots more crucial than regular-season game-winning shots?
Compare clutch performances from different times, teams, and systems. Data
collection and analysis biases may impact research results. Data from playoff
games may overrepresent postseason performers, biasing clutch performance.
The study must consider these biases and interpret findings nuancedly.
Addressing these issues will assure the findings' reliability and provide NBA
organizations and players useful data. By acknowledging dataset restrictions,
clutch performance difficulty, and interpretation challenges, NBA clutch
shooting research may improve.

### Summary Information

# Load necessary libraries
library(ggplot2)
library(dplyr)
library(readr)

# Load the CSV data
file_url <- "https://raw.githubusercontent.com/jai-agrawal/nba-player-clutch-stats/master/clutch_data.csv"
clutch_data <- read_csv(file_url)

# Calculate weighted statistics based on minutes played
clutch_data <- clutch_data %>%
  mutate(weighted_tov = TOV * MIN / GP,
         weighted_ast = AST * MIN / GP,
         weighted_stl = STL * MIN / GP,
         weighted_ft_pct = FT_PCT * MIN / GP,
         weighted_blk = BLK * MIN / GP)

# Filter data based on minimum minutes and games played
clutch_data <- clutch_data %>%
  filter(MIN >= 2, GP >= 15)

# Group data by player and calculate total weighted statistics
player_stats <- clutch_data %>%
  group_by(PLAYER_NAME) %>%
  summarise(total_weighted_tov = sum(weighted_tov, na.rm = TRUE),
            total_weighted_ast = sum(weighted_ast, na.rm = TRUE),
            total_weighted_stl = sum(weighted_stl, na.rm = TRUE),
            avg_weighted_ft_pct = mean(weighted_ft_pct, na.rm = TRUE),
            total_weighted_blk = sum(weighted_blk, na.rm = TRUE))

# Normalize each statistic
player_stats <- player_stats %>%
  mutate(norm_weighted_tov = (total_weighted_tov - min(total_weighted_tov)) / (max(total_weighted_tov) - min(total_weighted_tov)),
         norm_weighted_ast = (total_weighted_ast - min(total_weighted_ast)) / (max(total_weighted_ast) - min(total_weighted_ast)),
         norm_weighted_stl = (total_weighted_stl - min(total_weighted_stl)) / (max(total_weighted_stl) - min(total_weighted_stl)),
         norm_weighted_ft_pct = (avg_weighted_ft_pct - min(avg_weighted_ft_pct)) / (max(avg_weighted_ft_pct) - min(avg_weighted_ft_pct)),
         norm_weighted_blk = (total_weighted_blk - min(total_weighted_blk)) / (max(total_weighted_blk) - min(total_weighted_blk)))

# Calculate the composite score as the average of the normalized statistics
player_stats <- player_stats %>%
  rowwise() %>%
  mutate(composite_score = mean(c(norm_weighted_tov, norm_weighted_ast, norm_weighted_stl, norm_weighted_ft_pct, norm_weighted_blk), na.rm = TRUE)) %>%
  ungroup()

# Select the top 20 players based on the composite score
top_20_players <- player_stats %>%
  arrange(desc(composite_score)) %>%
  slice_head(n = 20)

# Select relevant columns to display
top_20_players <- top_20_players %>%
  select(PLAYER_NAME, total_weighted_tov, total_weighted_ast, total_weighted_stl, avg_weighted_ft_pct, total_weighted_blk, composite_score)



Our analysis of the clutch performance data from the 2020-21 NBA season
offers valuable insights into player performance during critical moments of
the game. We focused on key statistics, weighted by minutes played.

In turnovers (TOV), LeBron James had the highest weighted total of 0.148,
followed by James Harden and Kevin Durant. Turnovers indicate negative
output; thus, teams would want to avoid players who frequently turn the ball
over in clutch moments.

For assists (AST), LeBron James again led with a weighted total of 0.198,
with Jamal Murray and James Harden close behind. This highlights LeBron's
ability to facilitate plays under pressure.

In steals (STL), Collin Sexton topped the list with 0.06, followed by LeBron
James and Will Barton. Steals are crucial for defensive prowess in clutch
situations, showing a player's ability to disrupt opponents.

Jamal Murray had the highest average weighted free throw percentage (FT_PCT)
at 0.263, with Will Barton and Derrick White also performing well. This
indicates these players' reliability in scoring from free throws during
crucial moments.

Anthony Davis led in weighted blocks (BLK) with 0.0844, followed by Joel
Embiid and Robert Williams. Blocks are vital for defensive stances in clutch
times, highlighting these players' ability to protect the rim.

This detailed analysis identifies the most impactful players in clutch
situations, providing valuable insights into player performance under
pressure.


### Table
# Load necessary libraries
library(ggplot2)
library(dplyr)
library(readr)

# Load the CSV data
file_url <- "https://raw.githubusercontent.com/jai-agrawal/nba-player-clutch-stats/master/clutch_data.csv"
clutch_data <- read_csv(file_url)

# Calculate weighted statistics based on minutes played
clutch_data <- clutch_data %>%
  mutate(weighted_tov = TOV * MIN / GP,
         weighted_ast = AST * MIN / GP,
         weighted_stl = STL * MIN / GP,
         weighted_ft_pct = FT_PCT * MIN / GP,
         weighted_blk = BLK * MIN / GP)

# Filter data based on minimum minutes and games played
clutch_data <- clutch_data %>%
  filter(MIN >= 2, GP >= 15)

# Group data by player and calculate total weighted statistics
player_stats <- clutch_data %>%
  group_by(PLAYER_NAME) %>%
  summarise(total_weighted_tov = sum(weighted_tov, na.rm = TRUE),
            total_weighted_ast = sum(weighted_ast, na.rm = TRUE),
            total_weighted_stl = sum(weighted_stl, na.rm = TRUE),
            avg_weighted_ft_pct = mean(weighted_ft_pct, na.rm = TRUE),
            total_weighted_blk = sum(weighted_blk, na.rm = TRUE))

# Normalize each statistic
player_stats <- player_stats %>%
  mutate(norm_weighted_tov = (total_weighted_tov - min(total_weighted_tov)) / (max(total_weighted_tov) - min(total_weighted_tov)),
         norm_weighted_ast = (total_weighted_ast - min(total_weighted_ast)) / (max(total_weighted_ast) - min(total_weighted_ast)),
         norm_weighted_stl = (total_weighted_stl - min(total_weighted_stl)) / (max(total_weighted_stl) - min(total_weighted_stl)),
         norm_weighted_ft_pct = (avg_weighted_ft_pct - min(avg_weighted_ft_pct)) / (max(avg_weighted_ft_pct) - min(avg_weighted_ft_pct)),
         norm_weighted_blk = (total_weighted_blk - min(total_weighted_blk)) / (max(total_weighted_blk) - min(total_weighted_blk)))

# Calculate the composite score as the average of the normalized statistics
player_stats <- player_stats %>%
  rowwise() %>%
  mutate(composite_score = mean(c(norm_weighted_tov, norm_weighted_ast, norm_weighted_stl, norm_weighted_ft_pct, norm_weighted_blk), na.rm = TRUE)) %>%
  ungroup()

# Select the top 20 players based on the composite score
top_20_players <- player_stats %>%
  arrange(desc(composite_score)) %>%
  slice_head(n = 20)

# Select relevant columns to display
top_20_players <- top_20_players %>%
  select(PLAYER_NAME, total_weighted_tov, total_weighted_ast, total_weighted_stl, avg_weighted_ft_pct, total_weighted_blk, composite_score)

# Print the top 20 players
print(top_20_players)

This detailed analysis identifies the most impactful players in clutch
situations, offering valuable insights into their performance under pressure.
By focusing on weighted statistics, we can better understand which players
excel in critical moments and which areas require improvement.

### Chart 1

```{r, echo = FALSE, message = FALSE}
# Load necessary libraries
library(ggplot2)
library(dplyr)
library(readr)

# Load the CSV data
file_url <- "https://raw.githubusercontent.com/jai-agrawal/nba-player-clutch-stats/master/clutch_data.csv"
clutch_data <- read_csv(file_url)

# Calculate weighted statistics based on minutes played
clutch_data <- clutch_data %>%
  mutate(weighted_tov = TOV * MIN / GP)

# Filter data based on minimum minutes and games played
clutch_data <- clutch_data %>%
  filter(MIN >= 1, GP >= 3)

# Group data by team and calculate mean and standard deviation of turnovers
turnover_stats <- clutch_data %>%
  group_by(TEAM_ABBREVIATION) %>%
  summarise(avg_weighted_tov = mean(weighted_tov, na.rm = TRUE),
            sd_weighted_tov = sd(weighted_tov, na.rm = TRUE))

# Create bar chart with standard deviation
ggplot(turnover_stats, aes(x = TEAM_ABBREVIATION, y = avg_weighted_tov, fill = TEAM_ABBREVIATION)) +
  geom_bar(stat = "identity") +
  geom_errorbar(aes(ymin = avg_weighted_tov - sd_weighted_tov, ymax = avg_weighted_tov + sd_weighted_tov), width = 0.8, position = position_dodge(0.9)) +
  labs(title = "Average Turnovers in Clutch Time",
       x = "Team Abbreviation",
       y = "Average Turnovers") +
  theme_minimal() +
  theme(legend.position = "none")
```

I chose to generate a bar chart with error bars to illustrate the average
turnovers in clutch time for each NBA team alongside their standard
deviations. This chart effectively communicates both the average turnover
rates and the variability within each team, allowing for quick comparisons
between teams and insight into the consistency of performance. The height of
each bar indicates the average number of turnovers per team, while the error
bars represent the standard deviation, showing the spread or consistency of
turnover rates among players within each team. By examining the chart,
viewers can easily identify teams with higher or lower turnover rates during
crucial moments of the game and assess the stability of performance within
each team.

### Chart 2

```{r, echo = FALSE, message = FALSE}
# Load necessary libraries
library(ggplot2)
library(dplyr)
library(readr)

# Load the CSV data
file_url <- "https://raw.githubusercontent.com/jai-agrawal/nba-player-clutch-stats/master/clutch_data.csv"
clutch_data <- read_csv(file_url)

# Filter data based on minimum minutes and games played
clutch_data_filtered <- clutch_data %>%
  filter(MIN >= 1, GP >= 3)

# Calculate total minutes played for each player
clutch_data_filtered <- clutch_data_filtered %>%
  mutate(total_minutes = MIN * GP)

# Calculate weighted defensive metric
clutch_data_defensive <- clutch_data_filtered %>%
  mutate(defensive_metric = (STL + BLK) / total_minutes)

# Select top 20 players with highest defensive metric
top_20_defensive_players <- clutch_data_defensive %>%
  group_by(PLAYER_NAME) %>%
  summarise(total_defensive_metric = sum(defensive_metric, na.rm = TRUE)) %>%
  top_n(20, wt = total_defensive_metric)

# Merge top 20 players with original clutch data
clutch_data_top_20_defensive <- inner_join(clutch_data_defensive, top_20_defensive_players, by = "PLAYER_NAME")

# Create scatter plot of defensive metric vs FG_PCT for top 20 players
ggplot(clutch_data_top_20_defensive, aes(x = defensive_metric, y = FG_PCT, color = PLAYER_NAME)) +
  geom_point() +
  labs(title = "Defensive Metric vs Field Goal Percentage in Clutch Time (Top 20 Players by Defensive Metric)",
       x = "Defensive Metric",
       y = "Field Goal Percentage") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

I opted to construct a scatter plot depicting the relationship between the
defensive metric and the field goal percentage during clutch time for the top
20 players with the highest defensive metrics. This visualization effectively
communicates both the defensive performance and its impact on opponents'
shooting efficiency. Each point on the scatter plot represents a player, with
their defensive metric plotted on the x-axis and their field goal percentage
on the y-axis. By examining the chart, viewers can discern trends or patterns
in defensive effectiveness and its correlation with opponents' field goal
percentages. This visualization offers valuable insights into the defensive
prowess of the top players during critical moments of the game and sheds
light on their impact on opponent scoring efficiency.

### Chart 3

```r
# Load necessary libraries
library(ggplot2)
library(dplyr)
library(readr)

# Load the CSV data
file_url <- "https://raw.githubusercontent.com/jai-agrawal/nba-player-clutch-stats/master/clutch_data.csv"
clutch_data <- read_csv(file_url)

# Filter data based on minimum minutes and games played
clutch_data <- clutch_data %>%
  filter(MIN >= 1, GP >= 3)

# Group data by team and calculate average points scored per game and standard deviation
points_stats <- clutch_data %>%
  group_by(TEAM_ABBREVIATION) %>%
  summarise(avg_points = mean(PTS, na.rm = TRUE),
            sd_points = sd(PTS, na.rm = TRUE))

# Create bar chart of average points scored per game with error bars for standard deviation
ggplot(points_stats, aes(x = reorder(TEAM_ABBREVIATION, -avg_points), y = avg_points, fill = TEAM_ABBREVIATION)) +
  geom_bar(stat = "identity") +
  geom_errorbar(aes(ymin = avg_points - sd_points, ymax = avg_points + sd_points), width = 0.2, position = position_dodge(0.9)) +
  labs(title = "Average Points Scored per Game in Clutch Time by Team",
       x = "Team Abbreviation",
       y = "Average Points Scored") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")
```

I chose to generate a bar chart with error bars to illustrate the average
points scored per game in clutch time by each NBA team along with their
standard deviations. This chart effectively communicates both the average
scoring performance and the variability within each team's scoring during
crucial moments of the game. Each bar represents the average points scored
per game by a team, while the error bars indicate the standard deviation,
showing the spread or consistency of scoring performance among players within
each team. By examining the chart, viewers can easily identify teams with
higher or lower average scoring rates during clutch situations and assess the
stability of scoring performance within each team.
